Multivariate Neuroimage Analysis with \textit{ANTsR}: Application to Supervised Brain Tumor Segmentation using Random Forests

### 2014-01-17 paper.tex
WIP: BA comments

One of the innovations that we provide in this work is
the use of concatenated random forests for improved probabilistic estimation
of the brain segmentation labels.  Briefly, Gaussian mixture
modeling (GMM) is used to create a subset of feature images which are
used to train and predict during the first classification stage.  
The random forest voting output probabilistic maps are then used as 
spatial probability priors for performing a MAP-MRF segmentation
and subsequent feature image creation
thereby replacing the GMM-derived feature images for a second
classification stage.  This dual random forest model generation
and application
improves results and is detailed in later sections.

@@

In terms of implementation, this algorithm is 
encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
available in the ANTs repository, which permits parallel processing on
an individual workstation or on a large computational cluster.  In
Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
the Kirby multivariate symmetrical template consisting of FA, FLAIR,  
T1, and T2 components.  Although DTI-based images were not included in the
challenge data, such images have shown discriminative potential 
\citep[e.g.][]{price2003} to warrant investigation in our future work.

@@

We previously mentioned that all data was provided already 
skull-stripped.  However, we note to the reader that an
available ANTs-based solution to the brain extraction problem
is found in the script {\tt antsBrainExtraction.sh} which uses
a hybrid approach based on image registration and binary morphology
to isolate the brain \cite{avants2010a}.  We recently performed an evaluation of over 1200
MRI which demonstrated perfect performance \cite{tustison2013a}.

\paragraph{Stage 1 (GMM)}
Prior  cluster centers for specific tissue types learned from training data \citep{reynolds2009} are used in the GMM random forest stage to create 
multiple feature images.  
Given $M$ tissue types, a GMM formulates the 
probability distribution at each voxel, $\mathbf{x}$, as the
sum of Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
\begin{align}
p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
\end{align}
where $\sum_{m=1}^M \lambda_m = 1$.  One popular method for 
determining the parameters of the GMM is maximum likelihood 
estimation which can performed using the Atropos segmentation 
tool \citep{avants2011} available in ANTs.

@@

Prediction during the GMM stage results in a set of random 
forest voting classification maps giving a probabilistic estimate
of the label value at each voxel.  These spatial probability
maps can be used as spatial priors in a MAP-MRF segmentation
estimation protocol also performed using the Atropos segmentation 
tool \citep{avants2011} available in ANTs although we couple
it with iterative bias correction as contained in the script
{\tt antsAtroposN4.sh}.  The probabilistic output (and corresponding
feature images) 
for each modality are used to replace the analogous GMM-based feature images used 
during the first stage of prediction along with the other feature
images previously generated (such as the template-based data).
We discovered that this composite application of random
forest models improves the resulting labeled output.

@@

@@

@@

Although previous research has employed random forests for supervised brain
segmentation, our contribution of concatenated model application and use
of symmetric multivariate templates demonstrated good performance 
in the recent MICCAI 2013 Brain Tumor Segmentation challenge/workshop.  
In terms of variable importance, the latter provided several highly 
discriminative features which resulted in the top-performing algorithm 
of the competition.  This confirmed what others have found in that
random forests provide an excellent framework for prediction in certain
medical image analysis problems.

>>>>

One of the innovations that we provide in this work is
the use of \textit{concatenated random forests} \textcolor{red}{where
  does the name come from?  it doesnt get mentioned again until the
  discussion section ... expanding on this / defining it could help
  ``novelty'' angle in neuroimage submission .... FIXME} for improved probabilistic estimation
of the brain segmentation labels.  Briefly, Gaussian mixture
modeling (GMM) is used to create a subset of feature images which are
used to train and predict during the first classification stage.  
The random forest voting output probabilistic maps are then used as 
spatial probability priors for performing a MAP-MRF segmentation
and subsequent feature image creation
thereby replacing the GMM-derived feature images for a second
classification stage.  This dual random forest model generation
and application
improves results and is detailed in later sections.

@@

In terms of implementation, this algorithm is 
encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
available in the ANTs repository, which permits parallel processing on
an individual workstation or on a large computational cluster.  In
Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
the Kirby multivariate symmetrical template consisting of FA, FLAIR,  
T1, and T2 components.  Although DTI-based images were not included in the
challenge data, such images have shown discriminative potential 
\citep[e.g.][]{price2003} to warrant investigation in our future
work.  The general procedure for const ructing a symmetric template involves flipping data around the mid-sagittal
plane, as outlined in prior work
\citep{Hopkins2013}. \textcolor{red}{FIXME-nick, is that correct? is
  this covered in the scripts?  if so, where?}

@@

We previously mentioned that all data was provided already 
skull-stripped.  However, we note to the reader that an
available ANTs-based solution to the brain extraction problem
is found in the script {\tt antsBrainExtraction.sh} which uses
a hybrid approach based on image registration and binary morphology
to isolate the brain \cite{avants2010a}.  We recently performed an evaluation of over 1200
MRI which demonstrated performance that is comparable or better than
other available methodology \cite{tustison2013a}.

\paragraph{Stage 1 (GMM)} \textcolor{red}{Need more descriptive
  section name ... lots going on in the paper , easy to get disoriented}
Prior cluster centers for specific tissue types learned from training data \citep{reynolds2009} are used in the GMM random forest stage to create 
multiple feature images.  \textcolor{red}{FIXME I think this should be
  expanded upon a bit ... can you be more specific?  how are these
  cluster centers passed to the RF?  why are these centers portable to
  this new data? etcetera.}
Given $M$ tissue types, a GMM formulates the 
probability distribution at each voxel, $\mathbf{x}$, as the
sum of Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
\begin{align}
p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
\end{align}
where $\sum_{m=1}^M \lambda_m = 1$.  One popular method for 
determining the parameters of the GMM is maximum likelihood 
estimation which can performed using the Atropos segmentation 
tool \citep{avants2011} available in ANTs.  \textcolor{red}{FIXME
  would prefer active voice version of the sentence to make clear that
  Atropos uses EM} See Figure~\ref{fig:featureImages} for
an example segmentation.

\textcolor{red}{should there be an RF model equation here?  i.e. what
  exactly (in R) was used ?}

@@

Prediction during the GMM stage results in a set of random 
forest voting classification maps giving a probabilistic estimate
of the label value at each voxel.  \textcolor{red}{FIXME ---i think
  this sentence is confusing ... did you already run the RF at this
  point?  seems no ... but ``results in a set of RF voting
  classification maps'' sounds like you already have them.}  These spatial probability
maps can be used as spatial priors in a MAP-MRF segmentation
estimation protocol also performed using the Atropos segmentation 
tool \citep{avants2011} available in ANTs although we couple
it with iterative bias correction as contained in the script
{\tt antsAtroposN4.sh}.  The probabilistic output (and corresponding
feature images) 
for each modality are used to replace the analogous GMM-based feature images used 
during the first stage of prediction along with the other feature
images previously generated (such as the template-based data).
We discovered that this composite application of random
forest models improves the resulting labeled output.

\textcolor{red}{ok ... but i still don't understand this ... it gives
  me a vague idea ... can you sketch out the equations for these
  features/predictors? and give better intuition of what the
  ``composite'' RF (previously referred to as concatenated?) is?}

@@

\textcolor{red}{do you actually use variable importance or no?
  i.e. to trim the number of features?  overall, having a hard time
  seeing what is possible versus what was actually done ... }

@@

\textcolor{red}{how novel is this a/symmetry stuff?  i have the chimp
  paper that briefly mentions it but could this be spun as a
  ``contribution'' (with additional technical detail) in addition to pointing out the benefit of asymmetry
  plus RF's?  also see
  \href{http://stnava.github.io/asymmetry/}{http://stnava.github.io/asymmetry/}
  which is sort of novel ... }

@@

Although previous research has employed random forests for supervised brain
segmentation, our contribution of \textcolor{red}{concatenated model} application and use
of symmetric multivariate templates demonstrated good performance 
in the recent MICCAI 2013 Brain Tumor Segmentation challenge/workshop.  
In terms of variable importance, the latter provided several highly 
discriminative features which resulted in the top-performing algorithm 
of the competition.  This confirmed what others have found in that
random forests provide an excellent framework for prediction in certain
medical image analysis problems.  Even with relatively few training
subjects relative to input variables, the random forest models
perform generalizably.
In addition, this work highlights the value of  a symmetric multiple
modality template for clinical disorders in which asymmetry is a
hallmark.  Furthermore, our challenge-leading results establish the
value of such templates in multiple modality prediction
tasks in which there is a small training set with an abundance of
multivariate data.

\textcolor{red}{\subsection{Discussion of clinical relevance of the multivariate
  features we ``found''}}
